%% 
%% Copyright 2007-2018 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

\documentclass[preprint,10pt,number,5p,twocolumn]{elsarticle}
 
 
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

%\usepackage[acronym,smallcaps]{glossaries}

%Timeline
%\usepackage[utf8]{inputenc}
%\usepackage[TS1,T1]{fontenc}
%\usepackage{fourier, heuristica}
%\usepackage{array, booktabs}
%\usepackage{graphicx}
%\usepackage[x11names,table]{xcolor}
%\usepackage{caption}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[x11names,table,xcdraw]{xcolor}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage{makecell}
\usepackage{subfigure}
\usepackage{lipsum}
\usepackage{pifont}
\usepackage{tabularx}
\usepackage{ctable}
\usepackage{pbox}
\usepackage[T1,hyphens]{url}  
\usepackage{multirow}
\usepackage{makecell}
\usepackage{hhline}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage[acronym,smallcaps]{glossaries}
\usepackage[normalem]{ulem}
\useunder{\uline}{\ul}{}

\usepackage{array, booktabs}
\usepackage{graphicx}

\newcommand{\foo}{\color{LightSteelBlue3}\makebox[0pt]{\textbullet}\hskip-0.5pt\vrule width 1pt\hspace{\labelsep}}

\input{./acronyms}

\journal{Computer Communications}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%\title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{Network Service Orchestration: A Survey}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

\author[First]{Nathan F. Saraiva de Sousa}
\author[First]{Danny A. Lachos Perez}
\author[First]{Raphael V. Rosa}
\author[Second]{Mateus A. S. Santos}
\author[First]{Christian Esteve Rothenberg}

\address[First]{Department of Computer Engineering and Industrial Automation, School of Electrical and Computer Engineering, University of Campinas - UNICAMP, Campinas, SP, Brazil}
\address[Second]{Ericsson Research, Indaiatuba, SP, Brazil}

\fntext[*]{This work is under submission for peer review. Until then, and even after an eventual publication, the authors are most welcome for any feedback to improve the work and turn the \textit{github} and \textit{arxiv} versions of this publication a ``living document'' driven by community contributions as NSO evolves.  
Do not hesitate to contact the authors and/or submit \textit{github} pull requests or issues: 
https://github.com/intrig-unicamp/publications/tree/master/NSO-Survey.}

\begin{abstract}
Business models of network service providers are undergoing an evolving transformation fueled by vertical customer demands and technological advances such as 5G, Software Defined Networking~(SDN), and Network Function Virtualization~(NFV). Emerging scenarios call for agile network services consuming network, storage, and compute resources across heterogeneous infrastructures and administrative domains. Coordinating resource control and service creation across interconnected domains and diverse technologies becomes a grand  challenge. Research and development efforts are being devoted to enabling orchestration processes to  automate, coordinate, and manage the deployment and operation of network services. In this survey, we delve into the topic of Network Service Orchestration~(NSO) by reviewing the historical background, relevant research projects, enabling technologies, and standardization activities. We define key concepts and propose a taxonomy of NSO approaches and solutions to pave the way to a common understanding of the various ongoing efforts towards the realization of diverse NSO application scenarios. Based on the analysis of the state of affairs, we present a series of open challenges and research opportunities, altogether contributing to a timely and comprehensive survey on the vibrant and strategic topic of network service orchestration.

\end{abstract}

\begin{keyword}
Network Service Orchestration~(NSO) \sep SDN \sep NFV \sep multi-domain \sep orchestration \sep virtualization \sep lifecycle management
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}

%% \linenumbers

\input{TEXs/01_Introduction}
\input{TEXs/02_Background.tex}
\input{TEXs/03_Enabling_Scenarios.tex}
\input{TEXs/04_NSO.tex}
\input{TEXs/05_NSO_Standardization.tex}
\input{TEXs/06_Projects.tex}
\input{TEXs/07_Solutions.tex}
\input{TEXs/08_Challenges.tex}
\input{TEXs/09_Conclusions.tex}


%% main text
%\section{}
%\label{}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% If you have bibdatabase file and want bibtex to generate the
%% bibitems, please use
%%
%%  \bibliographystyle{elsarticle-harv} 
%%  \bibliography{<your bibdatabase>}

%% else use the following coding to input the bibitems directly in the
%% TeX file.

%\begin{thebibliography}{00}

%% \bibitem[Author(year)]{label}
%% Text of bibliographic item

%\bibitem[ ()]{}

%\end{thebibliography}

\section*{Acknowledgment}

This research was supported by the Innovation Center, Ericsson S.A., Brazil, grant UNI.62. The authors would also like to express their gratitude to review contributions from David Moura,  Lucian Beraldo, Nazrul Islam, and Suneet Singh (in alphabetical order) funded by the EU-Brazil NECOS project under grant agreement no. 777067.
The authors are thankful for any feedback to improve the work. Do not hesitate to contact the authors and/or via \textit{github}: 
https://github.com/intrig-unicamp/publications/tree/master/NSO-Survey.


\bibliographystyle{elsarticle-harv} 
\bibliography{bib/refs}

\end{document}

\endinput
%%
%% End of file `elsarticle-template-harv.tex'.
